Simple repression draft

### 2020-05-26 doc/appendix_03_regulated_promoters.tex
MRM pass through bursty promoter derivation on 20200525

In order to solve Eq.~\ref{eq:si_master_unreg} we will use the generating
function method \mrm{reference?}. The probability generating function is defined
as
F(z,t) = \sum_{m=0}^\infty z^m p(m,t),
where $z$ is just a dummy variable that will help us later on to obtain the
moments of the distribution. Applying $F(z, t)$ on both sides of 
Eq.~\ref{eq:si_master_unreg} results in
\begin{equation}
\sum_m z^m {d \over d\tau} p(m, \tau) = 
\sum_m z^m \left[ 
- m p(m, \tau) 
+ (m + 1) p(m + 1, \tau) 
+ \lambda \sum_{m' = 0}^m G_{m-m'} p(m', \tau) - \lambda p(m, \tau)
\end{equation}
where we use $\sum_m \equiv \sum_{m=0}^\infty$. We can distribute the sum and
use the definition of $F(z, t)$ to obtain
\begin{equation}
{d F(z, \tau) \over d\tau} =
- \sum_m z^m m p(m, \tau) +
+ \sum_m z^m (m + 1) p(m + 1, \tau)
+ \lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau)
- \lambda F(z, \tau).
\label{eq:si_generating_01}
\end{equation}
We can make use of properties of the generating function to write everything in
terms of $F(z, \tau)$: the first term on the right hand side of
Eq.~\ref{eq:si_generating_01} can be rewritten as
\sum_{m} z^{m} \cdot m \cdot p(m, \tau) &=
\sum_{m} z \frac{\partial z^{m}}{\partial z} p(m, \tau), \\
&=\sum_{m} z \frac{\partial}{\partial z}\left(z^{m} p(m, \tau)\right), \\
&=z \frac{\partial}{\partial z}\left(\sum_{m} z^{m} p(m, \tau)\right), \\
&= z {\partial F(z, \tau) \over \partial z}.
For the second term on the right hand side of Eq.~\ref{eq:si_generating_01} we
define $k \equiv m + 1$. This allows us to write
\sum_{m=0}^{\infty} z^{m} \cdot(m+1) \cdot p(m+1, \tau) &=
\sum_{k=1}^{\infty} z^{k-1} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=1}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=0}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \left(z \frac{\partial F(z)}{\partial z}\right), \\
&=\frac{\partial F(z)}{\partial z}.
The third term in Eq.~\ref{eq:si_generating_01} is the most trouble. The trick
is to reverse the default order of the sums as
\begin{equation}
\sum_{m=0}^{\infty} \sum_{m'=0}^{m} = \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty}.
\end{equation}
This results in
\lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau) =
\lambda \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} z^m 
\theta (1 - \theta)^{m-m'} p(m', \tau),
where we also substituted the definition of the geometric distribution $G_{k} =
\theta (1 - \theta)^k$. Redistributing the sums we can write
\lambda \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} z^m 
\theta (1 - \theta)^{m-m'} p(m', \tau) = 
\lambda \theta \sum_{n=0}^{\infty}(1-\theta)^{m'} P(m', \tau) 
\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m}.
\label{eq:si_generating_02}
We know that $1 - \theta \in (0, 1)$, and that $z$ is a dummy variable that we
usually evaluate as $z=1$. With these two facts we can safely state that
$| z (1 - \theta) | < 1$. Defining $n \equiv m - m'$ we rewrite the last sum in 
Eq.~\ref{eq:si_generating_02} as
\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m} &= 
\sum_{n=0}^{\infty} \left[z (1-\theta)\right]^{n + m'} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\sum_{n=0}^{\infty} \left[ z (1 - \theta) \right]^{m'} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\left( {1 \over 1 - z (1 - \theta)} \right),
where we use the geometric series since, as stated before,$| z (1 - \theta) | <
1$
Putting these results together, the PDE for the generation function is
\pderiv[F]{t} = \pderiv[F]{z} - z\pderiv[F]{z} - \lambda F
        + \frac{\lambda\theta F}{1-z(1-\theta)}.
Changing variables to $\xi=1-\theta$ and simplifying gives
\pderiv[F]{t} + (z - 1) \pderiv[F]{z} = \frac{(z-1)\xi}{1-z\xi}\lambda F.
\label{eq:1state_unreg_015}

@@

At steady-state, the PDE reduces to the ODE
\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
which we can integrate as
\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
The initial conditions for generating functions can be subtle and confusing.
The key fact follows from the definition
$F(z,t) = \sum_m z^m p(m,t)$,
normalization requires that
$F(z=1^-,t) = \sum_m p(m,t) = 1$.\footnote{
Sometimes the generating function may be undefined \textit{at} $z=1$ but the
limit still holds. Also many authors change variables from $z$ to other
things, so one must keep track of how the normalization condition transforms.
Doing the integrals (and producing constant $c$) gives
\ln F &= -\lambda \ln(1-\xi z) + c
F &= \frac{c}{(1-\xi z)^\lambda}.
Only one choice for $c$ can satisfy initial conditions, producing
F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
which is exactly the negative binomial's generating function, as expected.
Now we may recover the mRNA probability distribution of interest
by differentiating the generating function, or by consulting a
standard reference\mmnote{maybe van Kampen, check},
from which we find that
\begin{equation}
p(m) = \frac{\Gamma(m+\lambda)}{\Gamma(m+1)\Gamma(\lambda)}
        \theta^\lambda (1-\theta)^m.
\end{equation}
Note that the ratio of gamma functions is often expressed as a
binomial coefficient, but since $\lambda$ may be non-integer,
this would be ill-defined.
Re-expressing this exclusively in our variables of interest,
burst rate $\lambda$ and mean burst size $b$, we have
\begin{equation}
p(m) = \frac{\Gamma(m+\lambda)}{\Gamma(m+1)\Gamma(\lambda)}
        \left(\frac{1}{1+b}\right)^\lambda
        \left(\frac{b}{1+b}\right)^m.
\label{eq:nbinom_deriv_final}
\end{equation}

>>>>

In order to solve Eq.~\ref{eq:si_master_unreg} we will use the generating
function method \mrm{reference?}. The probability generating function is defined
as
\begin{align}
F(z,t) = \sum_{m=0}^\infty z^m p(m,t),
\end{align}
where $z$ is just a dummy variable that will help us later on to obtain the
moments of the distribution. Applying $F(z, t)$ on both sides of 
Eq.~\ref{eq:si_master_unreg} results in
\begin{equation}
\sum_m z^m {d \over d\tau} p(m, \tau) = 
\sum_m z^m \left[ 
- m p(m, \tau) 
+ (m + 1) p(m + 1, \tau) 
+ \lambda \sum_{m' = 0}^m G_{m-m'} p(m', \tau) - \lambda p(m, \tau)
\right],
\end{equation}
where we use $\sum_m \equiv \sum_{m=0}^\infty$. We can distribute the sum and
use the definition of $F(z, t)$ to obtain
\begin{equation}
{d F(z, \tau) \over d\tau} =
- \sum_m z^m m p(m, \tau) +
+ \sum_m z^m (m + 1) p(m + 1, \tau)
+ \lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau)
- \lambda F(z, \tau).
\label{eq:si_generating_01}
\end{equation}
We can make use of properties of the generating function to write everything in
terms of $F(z, \tau)$: the first term on the right hand side of
Eq.~\ref{eq:si_generating_01} can be rewritten as
\begin{align}
\sum_{m} z^{m} \cdot m \cdot p(m, \tau) &=
\sum_{m} z \frac{\partial z^{m}}{\partial z} p(m, \tau), \\
&=\sum_{m} z \frac{\partial}{\partial z}\left(z^{m} p(m, \tau)\right), \\
&=z \frac{\partial}{\partial z}\left(\sum_{m} z^{m} p(m, \tau)\right), \\
&= z {\partial F(z, \tau) \over \partial z}.
\end{align}
For the second term on the right hand side of Eq.~\ref{eq:si_generating_01} we
define $k \equiv m + 1$. This allows us to write
\begin{align}
\sum_{m=0}^{\infty} z^{m} \cdot(m+1) \cdot p(m+1, \tau) &=
\sum_{k=1}^{\infty} z^{k-1} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=1}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \sum_{k=0}^{\infty} z^{k} \cdot k \cdot p(k, \tau), \\
&=z^{-1} \left(z \frac{\partial F(z)}{\partial z}\right), \\
&=\frac{\partial F(z)}{\partial z}.
\end{align}
The third term in Eq.~\ref{eq:si_generating_01} is the most trouble. The trick
is to reverse the default order of the sums as
\begin{equation}
\sum_{m=0}^{\infty} \sum_{m'=0}^{m} = \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty}.
\end{equation}
This results in
\begin{align}
\lambda \sum_m z^m \sum_{m'=0}^m G_{m-m'} p(m', \tau) =
\lambda \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} z^m 
\theta (1 - \theta)^{m-m'} p(m', \tau),
\end{align}
where we also substituted the definition of the geometric distribution $G_{k} =
\theta (1 - \theta)^k$. Redistributing the sums we can write
\begin{align}
\lambda \sum_{m'=0}^{\infty} \sum_{m=m'}^{\infty} z^m 
\theta (1 - \theta)^{m-m'} p(m', \tau) = 
\lambda \theta \sum_{n=0}^{\infty}(1-\theta)^{m'} P(m', \tau) 
\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m}.
\label{eq:si_generating_02}
\end{align}
We know that $1 - \theta \in (0, 1)$, and that $z$ is a dummy variable that we
usually evaluate as $z=1$. With these two facts we can safely state that
$| z (1 - \theta) | < 1$. Defining $n \equiv m - m'$ we rewrite the last sum in 
Eq.~\ref{eq:si_generating_02} as
\begin{align}
\sum_{m=m'}^{\infty} \left[z (1-\theta)\right]^{m} &= 
\sum_{n=0}^{\infty} \left[z (1-\theta)\right]^{n + m'} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\sum_{n=0}^{\infty} \left[ z (1 - \theta) \right]^{m'} \\
&= \left[ z (1 - \theta) \right]^{m'} 
\left( {1 \over 1 - z (1 - \theta)} \right),
\end{align}
where we use the geometric series since, as stated before,$| z (1 - \theta) | <
1$. Putting these results together, the PDE for the generation function is
\begin{equation}
{\partial F \over \partial \tau} = 
{\partial F \over\partial z}
- z {\partial F \over \partial z} - \lambda F
+ \frac{\lambda\theta F}{1-z(1-\theta)}.
\end{equation}
Changing variables to $\xi=1-\theta$ and simplifying gives
\begin{align}
{\partial F \over \partial \tau} + (z - 1) {\partial F \over \partial z} = 
\frac{(z-1)\xi}{1-z\xi}\lambda F.
\label{eq:1state_unreg_015}
\end{align}

@@

To get at the mRNA distribution at steady state we first must solve
Eq.~\ref{eq:1state_unreg_015} setting the time derivative to zero. At
steady-state, the PDE reduces to the ODE
\begin{align}
\deriv[F]{z} = \frac{\xi}{1-z\xi}\lambda F,
\end{align}
which we can integrate as
\begin{align}
\int \frac{dF}{F} = \int \frac{\lambda\xi dz}{1-\xi z}.
\end{align}
The initial conditions for generating functions can be subtle and confusing. The
key fact follows from the definition
$F(z,t) = \sum_m z^m p(m,t)$,
normalization requires that
$F(z=1, t) = \sum_m p(m,t) = 1$.
\footnote{
Sometimes the generating function may be undefined \textit{at} $z=1$ but the
limit still holds. Also many authors change variables from $z$ to other
things, so one must keep track of how the normalization condition transforms.
}
Doing the integrals (and producing constant $c$) gives
\begin{align}
\ln F &= -\lambda \ln(1-\xi z) + c
\\
F &= \frac{c}{(1-\xi z)^\lambda}.
\end{align}
Only one choice for $c$ can satisfy initial conditions, producing
\begin{align}
F(z) = \left(\frac{1-\xi}{1-\xi z}\right)^\lambda
        = \left(\frac{\theta}{1 - z(1-\theta)}\right)^\lambda,
\end{align}
which is exactly the negative binomial's generating function. Now we may recover
the mRNA probability distribution of interest by differentiating the generating
function, or by consulting a standard reference\mmnote{maybe van Kampen, check},
from which we find that\mrm{I have a series of notes where I go through some 
of the derivatives to find the general pattern, arriving to this exact result 
by ``scratch and sniff'' without the black magic. Should I add it to this 
section?}
\begin{equation}
p(m) = \frac{\Gamma(m+\lambda)}{\Gamma(m+1)\Gamma(\lambda)}
        \theta^\lambda (1-\theta)^m.
\end{equation}
Note that the ratio of gamma functions is often expressed as a binomial
coefficient, but since $\lambda$ may be non-integer, this would be ill-defined.
Re-expressing this exclusively in our variables of interest, burst rate
$\lambda$ and mean burst size $b$, we have
\begin{equation}
p(m) = \frac{\Gamma(m+\lambda)}{\Gamma(m+1)\Gamma(\lambda)}
        \left(\frac{1}{1+b}\right)^\lambda
        \left(\frac{b}{1+b}\right)^m.
\label{eq:nbinom_deriv_final}
\end{equation}
